# 💻 Dummy Variables {#dummy-vars}


Categorical variables (also known as factor or qualitative variables) are variables that classify observations into groups. They have a limited number of different values, called levels. For example the gender of individuals are a categorical variable that can take two levels: Male or Female.

Regression analysis requires numerical variables. So, when a researcher wishes to include a categorical variable in a regression model, supplementary steps are required to make the results interpretable.


## examples data from `car` pkg `Salaries`

We’ll use the `Salaries` data set  from `car` pkg, which contains 2008/2009 nine-month academic salary for **Assistant Professors**, **Associate Professors** and **Professors** in a college in the U.S.


```{r, include=FALSE}
library(car)
data("Salaries")
Salaries = as.data.frame(Salaries)
```

```{r,echo=FALSE }
library(gt)

gt_preview(Salaries, top_n = 5, bottom_n = 1, incl_rownums = TRUE)
```

## Categorical variables with two levels

Since now we have pretty much seen regression with only numeric predictors, but the most of the times you are going to deal with categorical predictors like we did for regression in `iris` dataset.

Recall that, the regression equation, for predicting an outcome variable $Y$ on the basis of a predictor variable $X$, can be simply written as $Y = \alpha + \beta_1*X$. 

Suppose that, we wish to investigate differences in salaries between males and females.

Based on the gender variable, we can create a new dummy variable that takes the value:

- **1** if a person is male
- **0** if a person is female

and use this variable as a predictor in the regression equation, leading to the following the model:

- $\alpha + \beta_1$ if person is male
- $\alpha$ if person is female

The coefficients can be interpreted as follow:

- $\alpha$ is the average salary among females,
- $\alpha + \beta_1$ is the average salary among males,
- and $\beta_1$ is the average difference in salary between males and females.


For simple demonstration purpose, the following example models the salary difference between males and females by computing a simple linear regression model on the `Salaries` data set.
R creates dummy variables automatically:

```{r}
model <- lm(salary ~ sex, data = Salaries)
summary(model)$coef

```

From the output above, the average salary for female is estimated to be **101002**, whereas males are estimated a total of **101002** + **14088** = **115090**.

The p-value for the dummy variable sexMale is very significant, suggesting that there is a statistical evidence of a **difference in average salary between the genders.**


What happened is taht R has created a `sexMale` dummy variable that takes on a value of 1 if the sex is Male, and 0 otherwise (Females). The decision to code males as *1* and females as *0* (baseline) is arbitrary, and has no effect on the regression computation, but does alter the interpretation of the coefficients.

If you are willign to change the factor orders than you may expect to find a negative coefficient for `SexFemale.`


## Categorical variables with more than **two** levels

Generally, a categorical variable with $n$ levels will be transformed into n-1 variables each with two levels. These $n-1$ new variables contain the same information than the single variable. 

Lets take for example rank in the Salaries data has three levels: `AsstProf`, `AssocProf` and `Prof.` This variable could be dummy coded into two variables, one called `AssocProf` and one `Prof` (i.e. $n-1$)
That is to say:

- If rank = `AssocProf`, then the column `AssocProf` would be coded with a 1 and `Prof` with a 0.
- If rank = `Prof`, then the column `AssocProf` would be coded with a 0 and `Prof` would be coded with a 1.
- If rank = `AsstProf`, then both columns `AssocProf` and `Prof` would be coded with a 0.


This dummy coding is automatically performed by R. For demonstration purpose, you can use the function model.matrix() to create a contrast matrix for a factor variable, this is how it would look like (remember R handles that for you):


```{r}
res <- model.matrix(~rank, data = Salaries)
head(res[, -1])
```


When building linear model, there are different ways to encode categorical variables, known as contrast coding systems. The default option in R is to use the first level of the factor as a reference and interpret the remaining levels relative to this level.

Now let's fit the model and see results:

```{r}
library(car)
model2 <- lm(salary ~ yrs.service + rank + discipline + sex,
             data = Salaries)
summary(model2)
```

For example, it can be seen that being from discipline B (applied departments) is significantly associated with an average increase of 13473.38 in salary compared to discipline A (theoretical departments).

## oh there's another way to say this.. **OneHotEconding**

One-hot encoding is the process of converting a categorical variable with multiple categories into multiple variables, each with a value of 1 or 0. There are many packages that does that even if R handles that for you.

For the methods outlined below, the following simple dataframe will be required:

```{r}
set.seed(28)

data <- data.frame(
  Outcome = seq(1,100,by=1),
  Variable = sample(c("Red","Green","Blue"), 100, replace = TRUE)
)
```

Now starting from this dataframe you can one hot encode i.e. convert factors to 0s and 1s such as:


```{r}
library(caret)

dummy <- dummyVars(" ~ .", data=data)
newdata <- data.frame(predict(dummy, newdata = data)) 

```


## handling factors in R

Factors are used to represent categorical data. Factors can be ordered or unordered and are an important class for statistical analysis and for plotting.

Factors are stored as integers, and have labels associated with these unique integers. While factors look (and often behave) like character vectors, they are actually integers under the hood, and you need to be careful when treating them like strings.

Once created, factors can only contain a pre-defined set values, known as levels. By default, R always sorts levels in alphabetical order. For instance, if you have a factor with 2 levels:


```{r}
sex <- factor(c("male", "female", "female", "male"))
levels(sex)
```

```{r}
nlevels(sex)
```

now with 3

```{r}
food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(food)
```



## exercises


::: {.exercise #dv1}

Suppose you have a dataset with a categorical variable named "color" that has the following levels: "Red," "Blue," "Green," and "Yellow." Create a set of dummy variables for this categorical variable in R, making sure to use the appropriate method for encoding these levels.

:::

```{r}
# Create a sample dataset
your_dataset <- data.frame(
  color = c("Red", "Blue", "Green", "Yellow", "Red")
)
```



::: {.exercise #dv2}

Using the "Salaries" dataset, perform a linear regression analysis to investigate the effect of "discipline" (with levels "A" and "B") on salary. Interpret the coefficients for each level of "discipline."


:::


::: {.exercise #dv3}

Suppose you have a dataset with two categorical variables, "region" and "language," and you want to apply one-hot encoding to both variables. Provide R code to perform one-hot encoding for both variables simultaneously.


:::


```{r}
# Create a sample dataset
your_dataset <- data.frame(
  region = c("North", "South", "East", "West"),
  language = c("English", "Spanish", "French", "German")
)

```


::: {.exercise #dv4}

You have a dataset with an ordered factor variable "size" representing "Small," "Medium," and "Large" sizes. How can you change the order of levels to "Large," "Medium," and "Small"? Provide R code to modify the factor levels.

:::

```{r}
# Create a sample dataset
your_dataset <- data.frame(
  size = factor(c("Small", "Medium", "Large", "Small", "Large"), ordered = TRUE)
)

```


## solutions

::: {.answer data-latex=""}
**Answer to Question** \@ref(exr:dv1):

You can create dummy variables for the "color" variable using the `model.matrix` function as follows:

    
    # Create dummy variables for color
    color_dummies <- model.matrix(~ color - 1, data = your_dataset)
    colnames(color_dummies) <- levels(your_dataset$color)


:::



::: {.answer data-latex=""}
**Answer to Question** \@ref(exr:dv2):

You can perform the linear regression analysis and interpret the coefficients for the "discipline" levels as follows:

    # Fit the linear regression model
    model4 <- lm(salary ~ discipline, data = Salaries)
    summary(model4)
    
    # Interpretation of coefficients


:::


::: {.answer data-latex=""}
**Answer to Question** \@ref(exr:dv3):

You can apply one-hot encoding to multiple categorical variables simultaneously using the `dummyVars` function from the `caret` package as follows:

    
    # Load the necessary packages
    library(caret)
    
    # Apply one-hot encoding for "region" and "language"
    dummy <- dummyVars("~ region + language", data = your_dataset)
    newdata <- data.frame(predict(dummy, newdata = your_dataset))


:::


::: {.answer data-latex=""}
**Answer to Question** \@ref(exr:dv4):


You can change the order of levels in the ordered factor variable "size" as follows:

    # Change the order of factor levels
    your_dataset$size <- factor(your_dataset$size, levels = c("Large", "Medium", "Small"), ordered = TRUE)
    
This code will reorder the levels of the "size" factor variable as specified.

:::



