# 🔬 Advanced Statistical Methods

This chapter covers advanced statistical methods including regression with dummy variables, logistic regression, factor analysis, and cluster analysis.

## Learning Objectives

By the end of this chapter, you will be able to:

- Understand and implement regression with dummy variables
- Perform logistic regression for binary outcomes
- Conduct factor analysis to reduce dimensionality
- Apply cluster analysis techniques
- Interpret results from advanced statistical methods
- Use R for advanced statistical analysis

## Regression with Dummy Variables

Dummy variables (also called indicator variables) are binary variables (0/1) used to represent categorical data in regression models.

### Creating Dummy Variables

```{r dummy-variables-example, eval=FALSE}
# Load required packages
library(tidyverse)
library(fastDummies)

# Create sample data with categorical variable
set.seed(123)
n <- 100
education <- sample(c("High School", "Bachelor", "Master", "PhD"), n, replace = TRUE)
experience <- rnorm(n, mean = 5, sd = 2)
salary <- 30000 + 5000 * (education == "Bachelor") + 
          8000 * (education == "Master") + 
          12000 * (education == "PhD") + 
          2000 * experience + rnorm(n, mean = 0, sd = 3000)

# Create data frame
data <- data.frame(education = education, experience = experience, salary = salary)

# Create dummy variables
data_dummy <- dummy_cols(data, select_columns = "education", remove_first_dummy = TRUE)

# View the data
head(data_dummy)

# Fit regression with dummy variables
model <- lm(salary ~ experience + education_Bachelor + education_Master + education_PhD, 
            data = data_dummy)

# View model summary
summary(model)

# Alternative: R automatically creates dummy variables
model_auto <- lm(salary ~ experience + education, data = data)
summary(model_auto)
```

### Interpreting Dummy Variable Coefficients

- **Reference category**: The category not included in the model (usually the first alphabetically)
- **Coefficients**: Represent the difference from the reference category
- **Example**: If "High School" is the reference, the coefficient for "Bachelor" represents the additional salary for Bachelor's degree holders

## Logistic Regression

Logistic regression is used when the dependent variable is binary (0/1, Yes/No, Success/Failure).

### Binary Logistic Regression

```{r logistic-regression-example, eval=FALSE}
# Load required packages
library(tidyverse)
library(broom)

# Create sample data for logistic regression
set.seed(123)
n <- 200
age <- rnorm(n, mean = 35, sd = 10)
income <- rnorm(n, mean = 50000, sd = 15000)
education_years <- rnorm(n, mean = 16, sd = 3)

# Create binary outcome (loan approval)
log_odds <- -2 + 0.05 * age + 0.0001 * income + 0.2 * education_years
prob <- exp(log_odds) / (1 + exp(log_odds))
loan_approved <- rbinom(n, 1, prob)

# Create data frame
data <- data.frame(age = age, income = income, education_years = education_years, 
                   loan_approved = loan_approved)

# Fit logistic regression
model <- glm(loan_approved ~ age + income + education_years, 
             data = data, family = binomial())

# View model summary
summary(model)

# Extract coefficients and odds ratios
coef_summary <- tidy(model, exponentiate = TRUE)
print(coef_summary)

# Predict probabilities
data$predicted_prob <- predict(model, type = "response")

# Create visualization
ggplot(data, aes(x = income, y = loan_approved)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = TRUE) +
  labs(
    title = "Logistic Regression: Loan Approval vs Income",
    x = "Income",
    y = "Loan Approved (0/1)"
  ) +
  theme_minimal()
```

### Interpreting Logistic Regression Results

- **Coefficients**: Represent the change in log-odds
- **Odds Ratios**: e^(coefficient) represents the multiplicative change in odds
- **Probabilities**: Use the logistic function to convert log-odds to probabilities

## Factor Analysis

Factor analysis is used to identify underlying latent factors that explain the correlations among observed variables.

### Exploratory Factor Analysis (EFA)

```{r factor-analysis-example, eval=FALSE}
# Load required packages
library(tidyverse)
library(psych)
library(GPArotation)

# Create sample data with underlying factors
set.seed(123)
n <- 300

# Factor 1: Math ability
math1 <- rnorm(n, mean = 0, sd = 1)
math2 <- 0.8 * math1 + rnorm(n, mean = 0, sd = 0.6)
math3 <- 0.7 * math1 + rnorm(n, mean = 0, sd = 0.7)

# Factor 2: Verbal ability
verbal1 <- rnorm(n, mean = 0, sd = 1)
verbal2 <- 0.9 * verbal1 + rnorm(n, mean = 0, sd = 0.4)
verbal3 <- 0.8 * verbal1 + rnorm(n, mean = 0, sd = 0.6)

# Create data frame
data <- data.frame(
  math_test1 = math1,
  math_test2 = math2,
  math_test3 = math3,
  verbal_test1 = verbal1,
  verbal_test2 = verbal2,
  verbal_test3 = verbal3
)

# Perform factor analysis
# First, check if data is suitable for factor analysis
cortest.bartlett(data)

# Determine number of factors
fa.parallel(data, fa = "fa", n.iter = 100)

# Perform factor analysis
fa_result <- fa(data, nfactors = 2, rotate = "varimax")
print(fa_result)

# Plot factor loadings
fa.diagram(fa_result)

# Extract factor scores
factor_scores <- factor.scores(data, fa_result)
data$factor1 <- factor_scores$scores[, 1]
data$factor2 <- factor_scores$scores[, 2]

# Visualize factor scores
ggplot(data, aes(x = factor1, y = factor2)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Factor Scores",
    x = "Factor 1 (Math Ability)",
    y = "Factor 2 (Verbal Ability)"
  ) +
  theme_minimal()
```

## Cluster Analysis

Cluster analysis groups similar observations together based on their characteristics.

### K-Means Clustering

```{r kmeans-clustering-example, eval=FALSE}
# Load required packages
library(tidyverse)
library(cluster)
library(factoextra)

# Create sample data with clusters
set.seed(123)
n <- 200

# Generate three clusters
cluster1 <- data.frame(
  x = rnorm(n/3, mean = 2, sd = 0.5),
  y = rnorm(n/3, mean = 2, sd = 0.5),
  cluster = 1
)

cluster2 <- data.frame(
  x = rnorm(n/3, mean = 6, sd = 0.5),
  y = rnorm(n/3, mean = 2, sd = 0.5),
  cluster = 2
)

cluster3 <- data.frame(
  x = rnorm(n/3, mean = 4, sd = 0.5),
  y = rnorm(n/3, mean = 6, sd = 0.5),
  cluster = 3
)

# Combine clusters
data <- rbind(cluster1, cluster2, cluster3)
data$cluster <- as.factor(data$cluster)

# Perform K-means clustering
kmeans_result <- kmeans(data[, 1:2], centers = 3, nstart = 25)
data$kmeans_cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters
ggplot(data, aes(x = x, y = y, color = cluster)) +
  geom_point(size = 2) +
  labs(
    title = "True Clusters",
    x = "X",
    y = "Y"
  ) +
  theme_minimal()

ggplot(data, aes(x = x, y = y, color = kmeans_cluster)) +
  geom_point(size = 2) +
  labs(
    title = "K-means Clusters",
    x = "X",
    y = "Y"
  ) +
  theme_minimal()

# Determine optimal number of clusters
fviz_nbclust(data[, 1:2], kmeans, method = "wss")
fviz_nbclust(data[, 1:2], kmeans, method = "silhouette")
```

### Hierarchical Clustering

```{r hierarchical-clustering-example, eval=FALSE}
# Perform hierarchical clustering
dist_matrix <- dist(data[, 1:2])
hclust_result <- hclust(dist_matrix, method = "ward.D2")

# Plot dendrogram
plot(hclust_result, main = "Hierarchical Clustering Dendrogram")

# Cut tree to get clusters
hclust_clusters <- cutree(hclust_result, k = 3)
data$hclust_cluster <- as.factor(hclust_clusters)

# Visualize hierarchical clusters
ggplot(data, aes(x = x, y = y, color = hclust_cluster)) +
  geom_point(size = 2) +
  labs(
    title = "Hierarchical Clusters",
    x = "X",
    y = "Y"
  ) +
  theme_minimal()
```

## Model Selection and Validation

### Cross-Validation for Logistic Regression

```{r cross-validation-example, eval=FALSE}
# Load required packages
library(tidyverse)
library(caret)

# Create sample data
set.seed(123)
n <- 300
x1 <- rnorm(n, mean = 0, sd = 1)
x2 <- rnorm(n, mean = 0, sd = 1)
x3 <- rnorm(n, mean = 0, sd = 1)

# Create binary outcome
log_odds <- -1 + 0.5 * x1 + 0.3 * x2 - 0.2 * x3
prob <- exp(log_odds) / (1 + exp(log_odds))
y <- rbinom(n, 1, prob)

# Create data frame
data <- data.frame(x1 = x1, x2 = x2, x3 = x3, y = y)

# Set up cross-validation
ctrl <- trainControl(method = "cv", number = 10, classProbs = TRUE)

# Train logistic regression model
model <- train(as.factor(y) ~ x1 + x2 + x3, 
               data = data, 
               method = "glm", 
               family = "binomial",
               trControl = ctrl)

# View results
print(model)
print(model$results)
```

## Best Practices

1. **Check assumptions** for each method
2. **Use appropriate sample sizes** for reliable results
3. **Validate models** using cross-validation
4. **Interpret results** in context of the research question
5. **Consider multiple methods** when appropriate
6. **Document decisions** and rationale

## Common Pitfalls

1. **Overfitting**: Including too many variables
2. **Multicollinearity**: Highly correlated predictors
3. **Sample size**: Insufficient data for reliable results
4. **Assumptions**: Not checking method-specific assumptions
5. **Interpretation**: Misunderstanding coefficients and results

## Summary

Advanced statistical methods provide powerful tools for analyzing complex data. Key points:

- Choose appropriate methods based on research questions
- Check assumptions and validate models
- Interpret results carefully and in context
- Use multiple methods when appropriate
- Document all decisions and rationale

## References

- Slides: Available in the course drive
- Additional resources and examples provided in class
- R documentation for specific packages used
