# 💻 Lab: PCA & Correspondence Analysis {#lab-pca-ca}

> **⚠️ Note on Slides**: If you encounter broken links in the PDF slides (`module_2_PCA_&_CA.pdf`), please refer to the "Additional Resources" section at the end of this lab for updated and working links.

This is a hands-on laboratory session to practice Principal Component Analysis (PCA) and Correspondence Analysis (CA) after the theoretical lecture. Work through these exercises step by step, and don't hesitate to ask questions!

## Setup

First, load the required packages:

```{r setup-lab-pca-ca, message=FALSE, warning=FALSE}
library(FactoMineR)  # For PCA and CA
library(factoextra)  # For visualization
library(corrplot)    # For correlation plots
library(dplyr)       # For data manipulation
```

## Part 1: Principal Component Analysis (PCA)

### Exercise 1.1: Understanding the Data - Wine Dataset

We'll work with the wine dataset from the `gclus` package, which contains chemical analyses of wines from three different cultivars.

```{r ex1-1-load-data}
# Load the wine dataset
library(gclus)
data(wine)

# Examine the structure
str(wine)
head(wine)
dim(wine)

# Check for missing values
sum(is.na(wine))
```

**Questions:**
1. How many observations and variables are in the dataset?
2. What types of variables do we have?
3. Are there any missing values?

### Exercise 1.2: Data Exploration and Preprocessing

Before performing PCA, let's explore the data and understand why standardization is important.

```{r ex1-2-explore}
# Calculate means for each variable
apply(wine, 2, mean)

# Calculate variances for each variable
apply(wine, 2, var)

# Calculate standard deviations
apply(wine, 2, sd)

# Create a correlation matrix
cor_matrix <- cor(wine)
cor_matrix

# Visualize correlations
corrplot(cor_matrix, method = "circle", type = "upper", 
         order = "hclust", tl.cex = 0.8)
```

**Questions:**
1. Do the variables have similar scales? Why is this important for PCA?
2. Which variables are highly correlated? What does this suggest?
3. Should we standardize the data? Why or why not?

### Exercise 1.3: Perform PCA

Now let's perform PCA on the wine dataset.

```{r ex1-3-pca}
# Perform PCA with standardization
res.pca.wine <- PCA(wine, scale.unit = TRUE, graph = FALSE)

# Print summary
print(res.pca.wine)
```

**Questions:**
1. How many principal components were created?
2. Why is this number equal to the number of variables (or n-1)?

### Exercise 1.4: Examine Eigenvalues

```{r ex1-4-eigenvalues}
# Get eigenvalues
eigenvalues <- get_eigenvalue(res.pca.wine)
eigenvalues

# Alternative way
res.pca.wine$eig

# Plot scree plot
fviz_eig(res.pca.wine, addlabels = TRUE, ylim = c(0, 50))
```

**Questions:**
1. What is the total inertia? What does it represent?
2. How much variance is explained by the first principal component?
3. How many components would you retain based on:
   - Kaiser's criterion (eigenvalue > 1)?
   - The scree plot (elbow method)?
   - Cumulative variance > 80%?

### Exercise 1.5: Analyze Variables

```{r ex1-5-variables}
# Variable information
res.pca.wine$var

# Correlations between variables and components
res.pca.wine$var$cor

# Coordinates of variables
res.pca.wine$var$coord

# Quality of representation (cos²)
res.pca.wine$var$cos2

# Contributions
res.pca.wine$var$contrib
```

**Questions:**
1. Which variables are most correlated with the first principal component?
2. Which variables contribute most to the first dimension?
3. Which variables are well represented in the first two dimensions (cos² > 0.5)?

### Exercise 1.6: Visualize Variables

```{r ex1-6-plot-vars, fig.width=10, fig.height=8}
# Basic variable plot
fviz_pca_var(res.pca.wine, col.var = "black")

# Variables colored by quality of representation
fviz_pca_var(res.pca.wine, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)

# Variables colored by contributions
fviz_pca_var(res.pca.wine, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

**Questions:**
1. What do the arrows represent in the variable plot?
2. What does the length of an arrow indicate?
3. What does the angle between two arrows tell you about the correlation between variables?

### Exercise 1.7: Analyze Individuals

```{r ex1-7-individuals}
# Individual information
res.pca.wine$ind

# Coordinates of individuals
head(res.pca.wine$ind$coord)

# Quality of representation
head(res.pca.wine$ind$cos2)

# Contributions
head(res.pca.wine$ind$contrib)
```

### Exercise 1.8: Visualize Individuals

```{r ex1-8-plot-ind, fig.width=10, fig.height=8}
# Basic individual plot
fviz_pca_ind(res.pca.wine, col.ind = "black")

# Individuals colored by cultivar (if available)
# First, check if cultivar information exists
if("class" %in% colnames(wine)) {
  fviz_pca_ind(res.pca.wine, 
               col.ind = wine$class,
               palette = c("#00AFBB", "#E7B800", "#FC4E07"),
               addEllipses = TRUE,
               legend.title = "Cultivar")
}

# Individuals colored by quality of representation
fviz_pca_ind(res.pca.wine, col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

**Questions:**
1. Do the wines cluster by cultivar in the PCA space?
2. Which individuals are outliers or poorly represented?

### Exercise 1.9: Biplot

```{r ex1-9-biplot, fig.width=10, fig.height=8}
# Create biplot
fviz_pca_biplot(res.pca.wine, 
                col.ind = if("class" %in% colnames(wine)) wine$class else "black",
                col.var = "#2E9FDF",
                repel = TRUE,
                legend.title = list(col = "Cultivar"))
```

**Questions:**
1. What does the biplot show?
2. How can you interpret the relationship between wines and chemical properties?

### Exercise 1.10: Dimension Description

```{r ex1-10-dimdesc}
# Describe dimensions
dimdesc(res.pca.wine, axes = 1:2, proba = 0.05)
```

**Questions:**
1. Which variables are significantly associated with dimension 1?
2. Which variables are significantly associated with dimension 2?
3. How would you interpret these dimensions?

---

## Part 2: Correspondence Analysis (CA)

### Exercise 2.1: Understanding CA Data - Hair and Eye Color

We'll work with a contingency table showing the relationship between hair color and eye color.

```{r ex2-1-create-data}
# Create a contingency table
# Hair color (rows) vs Eye color (columns)
hair_eye <- matrix(c(32, 11, 10, 3,    # Black hair
                     38, 50, 25, 15,    # Brown hair
                     10, 10, 7, 7,      # Red hair
                     3, 30, 5, 8),      # Blond hair
                   nrow = 4, byrow = TRUE)

rownames(hair_eye) <- c("Black", "Brown", "Red", "Blond")
colnames(hair_eye) <- c("Brown", "Blue", "Hazel", "Green")

# Convert to data frame for better display
hair_eye_df <- as.data.frame(hair_eye)
hair_eye_df

# Total
sum(hair_eye)
```

**Questions:**
1. What does this contingency table represent?
2. What is the total number of observations?
3. Are there any obvious associations between hair and eye color?

### Exercise 2.2: Perform Correspondence Analysis

```{r ex2-2-ca}
# Perform CA
res.ca <- CA(hair_eye, graph = FALSE)

# Print summary
print(res.ca)
```

**Questions:**
1. How many dimensions does CA create?
2. What is the total inertia? What does it represent?

### Exercise 2.3: Examine Eigenvalues and Inertia

```{r ex2-3-eigenvalues}
# Get eigenvalues
res.ca$eig

# Plot scree plot
fviz_screeplot(res.ca, addlabels = TRUE)

# Calculate percentage of inertia explained
inertia_percent <- res.ca$eig[,2]
inertia_percent
```

**Questions:**
1. How much inertia is explained by the first dimension?
2. How many dimensions would you retain? Why?

### Exercise 2.4: Analyze Rows (Hair Colors)

```{r ex2-4-rows}
# Row information
res.ca$row

# Row coordinates
res.ca$row$coord

# Row contributions
res.ca$row$contrib

# Row cos² (quality of representation)
res.ca$row$cos2
```

**Questions:**
1. Which hair color is best represented in the first dimension?
2. Which hair color contributes most to dimension 1?

### Exercise 2.5: Analyze Columns (Eye Colors)

```{r ex2-5-cols}
# Column information
res.ca$col

# Column coordinates
res.ca$col$coord

# Column contributions
res.ca$col$contrib

# Column cos²
res.ca$col$cos2
```

**Questions:**
1. Which eye color is best represented in the first dimension?
2. Which eye color contributes most to dimension 1?

### Exercise 2.6: Visualize CA Results

```{r ex2-6-plot-ca, fig.width=10, fig.height=8}
# Symmetric plot (default)
fviz_ca_biplot(res.ca, repel = TRUE)

# Asymmetric plot (rows in principal coordinates)
fviz_ca_biplot(res.ca, map = "rowprincipal", repel = TRUE)

# Asymmetric plot (columns in principal coordinates)
fviz_ca_biplot(res.ca, map = "colprincipal", repel = TRUE)
```

**Questions:**
1. What does proximity between a row point and a column point indicate?
2. Which hair colors are associated with which eye colors?
3. What does the distance from the origin tell you?

### Exercise 2.7: Chi-Square Test

```{r ex2-7-chisq}
# Perform chi-square test
chisq.test(hair_eye)

# CA provides chi-square statistics
summary(res.ca)
```

**Questions:**
1. Is there a significant association between hair and eye color?
2. What does the chi-square statistic tell us?

### Exercise 2.8: Real Dataset - Smoking and Exercise

Let's work with a more realistic example. Create a contingency table from survey data.

```{r ex2-8-real-data}
# Create a contingency table: Smoking status vs Exercise frequency
smoking_exercise <- matrix(c(50, 30, 20,    # Non-smoker
                             20, 25, 15,    # Occasional smoker
                             10, 15, 25),   # Regular smoker
                           nrow = 3, byrow = TRUE)

rownames(smoking_exercise) <- c("Non-smoker", "Occasional", "Regular")
colnames(smoking_exercise) <- c("High", "Moderate", "Low")

smoking_exercise_df <- as.data.frame(smoking_exercise)
smoking_exercise_df

# Perform CA
res.ca2 <- CA(smoking_exercise, graph = FALSE)
print(res.ca2)
```

### Exercise 2.9: Visualize and Interpret

```{r ex2-9-plot-ca2, fig.width=10, fig.height=8}
# Create biplot
fviz_ca_biplot(res.ca2, repel = TRUE,
               title = "CA: Smoking Status vs Exercise Frequency")

# Get detailed information
summary(res.ca2)
```

**Questions:**
1. What associations do you see between smoking and exercise?
2. How would you interpret the dimensions?
3. What insights can you draw from this analysis?

---

## Part 3: Integrated Exercise

### Exercise 3.1: Complete Analysis - USArrests Dataset

Perform a complete PCA analysis on the `USArrests` dataset and answer the following questions:

```{r ex3-1-complete}
# Load data
data(USArrests)
head(USArrests)

# Your tasks:
# 1. Explore the data (means, variances, correlations)
# 2. Perform PCA
# 3. Determine how many components to retain
# 4. Interpret the results
# 5. Create visualizations
# 6. Describe what each dimension represents

# Write your code here:
```

**Questions to Answer:**
1. Why is standardization important for this dataset?
2. How many principal components should be retained? Justify your answer.
3. What does the first principal component represent?
4. Which states are outliers or unusual?
5. Can you identify regional patterns in the data?

### Exercise 3.2: CA on Survey Data

Create your own contingency table (e.g., from a survey or research question) and perform CA.

```{r ex3-2-custom-ca}
# Create your contingency table here
# Example: Education level vs Job satisfaction
# my_table <- matrix(c(...), nrow = ..., ncol = ...)

# Perform CA
# res.ca.custom <- CA(my_table, graph = FALSE)

# Visualize and interpret
# fviz_ca_biplot(res.ca.custom, repel = TRUE)
```

**Questions:**
1. What is your research question?
2. What associations do you find?
3. How would you interpret the dimensions?

---

## Summary and Key Takeaways

After completing these exercises, you should be able to:

✅ **PCA:**
- Understand when and why to use PCA
- Perform PCA using FactoMineR
- Interpret eigenvalues, loadings, and contributions
- Determine the optimal number of components
- Visualize and interpret PCA results
- Understand the difference between covariance and correlation-based PCA

✅ **CA:**
- Understand when to use CA (categorical data in contingency tables)
- Perform CA using FactoMineR
- Interpret row and column coordinates
- Understand the relationship between CA and chi-square test
- Visualize and interpret CA results
- Understand the difference between symmetric and asymmetric plots

### Common Mistakes to Avoid

1. **Forgetting to standardize** in PCA when variables are on different scales
2. **Using PCA on categorical data** - use CA instead
3. **Over-interpreting dimensions** - not all dimensions are meaningful
4. **Ignoring quality of representation** - check cos² values
5. **Not checking assumptions** - understand your data before analysis

### Next Steps

- Practice with different datasets
- Try combining PCA with clustering (see clustering chapters)
- Explore other dimensionality reduction techniques
- Read about Multiple Correspondence Analysis (MCA) for multiple categorical variables

---

## Additional Resources

### R Documentation
- FactoMineR documentation: `?PCA`, `?CA`
- factoextra documentation: `?fviz_pca_var`, `?fviz_ca_biplot`

### Online Resources (All Links Verified and Working)

#### Official Package Documentation
- **FactoMineR CRAN**: https://cran.r-project.org/package=FactoMineR
- **FactoMineR GitHub**: https://github.com/husson/FactoMineR
- **FactoMineR Website**: http://factominer.free.fr/
- **factoextra CRAN**: https://cran.r-project.org/package=factoextra

#### Tutorials and Guides
- **STHDA PCA Guide**: http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/
- **STHDA CA Guide**: http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/113-ca-correspondence-analysis-in-r-the-essentials/
- **R Documentation Search**: https://www.rdocumentation.org/ (search for PCA, CA functions)

#### R Markdown Resources (from slides)
- **R Markdown Cheat Sheet**: https://www.rstudio.com/resources/cheatsheets/ (download R Markdown cheat sheet)
- **R Markdown Guide**: https://bookdown.org/yihui/rmarkdown/
- **RStudio Download**: https://www.rstudio.com/products/rstudio/download/
- **R Markdown Tutorial**: https://rmarkdown.rstudio.com/lesson-1.html

### Further Reading

- Review the theoretical slides for deeper understanding
- Check the main PCA chapter (Chapter 17) for more examples
- **Principal Component Analysis** by Jolliffe (textbook)
- **Correspondence Analysis in Practice** by Greenacre

**⚠️ Important**: If you find broken links in the PDF slides (`module_2_PCA_&_CA.pdf`), all the links above are verified and working. Use this section as your reference.

**Good luck with your analysis! 🍀**

